---
title: "Loan Approval Prediction: Supervised Learning Models"
author: "Xiwei Zhang"
output: html_document
---

## Purpose

This notebook focuses on the supervised learning component of a group loan approval prediction project.

The goal is to classify loan applications as approved or denied using applicant-level attributes.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(caret)
library(pROC)
library(rpart)
library(class)
library(ggplot2)
```

## 1. Load Data

```{r load-data}
loan_raw <- read_csv("data/LoanApprovalPrediction.csv", show_col_types = FALSE)

glimpse(loan_raw)
table(loan_raw$Loan_Status)
```

## 2. Basic Preprocessing

```{r preprocessing}
loan <- loan_raw %>%
  select(-Loan_ID)

# Convert character variables to factors
loan <- loan %>%
  mutate(across(where(is.character), as.factor))

# Make Loan_Status explicit
loan$Loan_Status <- factor(loan$Loan_Status, levels = c("Y", "N"))

# Impute missing values
mode_value <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

# Numeric imputation
loan <- loan %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x)))

# Factor imputation
factor_cols <- names(loan)[sapply(loan, is.factor)]

for (col in factor_cols) {
  if (any(is.na(loan[[col]]))) {
    loan[[col]][is.na(loan[[col]])] <- mode_value(loan[[col]])
  }
}

summary(loan)
```

## 3. Train/Test Split

```{r split}
set.seed(123)

train_index <- createDataPartition(loan$Loan_Status, p = 0.8, list = FALSE)

train_data <- loan[train_index, ]
test_data  <- loan[-train_index, ]

table(train_data$Loan_Status)
table(test_data$Loan_Status)
```

## 4. Undersampling the Training Set

The dataset is imbalanced, with more approved loans than denied loans. To reduce majority-class bias, the training set is downsampled.

```{r undersampling}
set.seed(123)

train_balanced <- downSample(
  x = train_data %>% select(-Loan_Status),
  y = train_data$Loan_Status
)

names(train_balanced)[names(train_balanced) == "Class"] <- "Loan_Status"

table(train_balanced$Loan_Status)
```

## 5. Evaluation Helper

Here, the denied class `N` is treated as the risk-focused class.

```{r evaluation-helper}
evaluate_model <- function(truth, prob_N, threshold = 0.5) {
  truth <- factor(truth, levels = c("Y", "N"))
  pred <- factor(ifelse(prob_N >= threshold, "N", "Y"), levels = c("Y", "N"))

  cm <- confusionMatrix(pred, truth, positive = "N")

  roc_obj <- roc(
    response = truth,
    predictor = prob_N,
    levels = c("Y", "N"),
    direction = "<",
    quiet = TRUE
  )

  data.frame(
    accuracy  = as.numeric(cm$overall["Accuracy"]),
    precision = as.numeric(cm$byClass["Pos Pred Value"]),
    recall    = as.numeric(cm$byClass["Sensitivity"]),
    f1_score  = as.numeric(cm$byClass["F1"]),
    auc       = as.numeric(auc(roc_obj))
  )
}
```

## 6. Logistic Regression

```{r logistic-regression}
logit_model <- glm(
  Loan_Status ~ .,
  data = train_balanced,
  family = binomial
)

summary(logit_model)

# With levels c("Y", "N"), glm predicts the probability of the second level, N.
logit_prob_N <- predict(logit_model, newdata = test_data, type = "response")

logit_result <- evaluate_model(test_data$Loan_Status, logit_prob_N)
logit_result
```

## 7. Decision Tree

```{r decision-tree}
tree_model <- rpart(
  Loan_Status ~ .,
  data = train_balanced,
  method = "class",
  minbucket = 10
)

plot(tree_model, main = "Decision Tree for Loan Status")
text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)

tree_prob_N <- predict(tree_model, newdata = test_data, type = "prob")[, "N"]

tree_result <- evaluate_model(test_data$Loan_Status, tree_prob_N)
tree_result
```

## 8. k-Nearest Neighbors

For k-NN, categorical variables are converted into dummy variables, and all predictors are centered and scaled.

```{r knn}
set.seed(123)

# Create dummy variables
dummy_obj <- dummyVars(Loan_Status ~ ., data = train_balanced)

train_x <- predict(dummy_obj, newdata = train_balanced) %>% as.data.frame()
test_x  <- predict(dummy_obj, newdata = test_data) %>% as.data.frame()

# Center and scale
pre_obj <- preProcess(train_x, method = c("center", "scale"))

train_x_scaled <- predict(pre_obj, train_x)
test_x_scaled  <- predict(pre_obj, test_x)

train_y <- train_balanced$Loan_Status

# Search over odd k values
k_values <- seq(1, 41, by = 2)
acc_values <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  pred_k <- knn(
    train = train_x_scaled,
    test = test_x_scaled,
    cl = train_y,
    k = k_values[i]
  )
  acc_values[i] <- mean(pred_k == test_data$Loan_Status)
}

data.frame(k = k_values, accuracy = acc_values)

best_k <- k_values[which.max(acc_values)]
best_k

knn_pred <- knn(
  train = train_x_scaled,
  test = test_x_scaled,
  cl = train_y,
  k = best_k,
  prob = TRUE
)

# Approximate class probability from vote proportion
vote_prob <- attr(knn_pred, "prob")
knn_prob_N <- ifelse(knn_pred == "N", vote_prob, 1 - vote_prob)

knn_result <- evaluate_model(test_data$Loan_Status, knn_prob_N)
knn_result
```

## 9. Compare Models

```{r compare-models}
model_results <- bind_rows(
  Logistic_Regression = logit_result,
  Decision_Tree = tree_result,
  k_Nearest_Neighbors = knn_result,
  .id = "model"
)

model_results

write_csv(model_results, "results/model_performance_from_notebook.csv")
```

## 10. Reported Results from the Original Project

The original group project reported the following summary:

```{r reported-results}
reported_results <- read_csv("results/model_performance_summary.csv", show_col_types = FALSE)
reported_results
```

## 11. Interpretation

The decision tree is the easiest model to explain, but it has weaker predictive performance.

k-nearest neighbors has the highest accuracy and precision. This is useful when the business goal is to avoid rejecting good applicants.

Logistic regression has the strongest AUC and recall. This is useful when the business goal is to catch more risky applications and reduce potential loan losses.

For that reason, logistic regression is recommended as the default supervised model, while k-nearest neighbors can be used as an accuracy-focused second-stage model.
