---
title: "Final Project"
author: "Lening Gao"
date: "2025-04-05"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

In financial services, accurately predicting loan approvals is essential
for mitigating risk and ensuring equitable lending practices. This
project aims to develop a predictive model for loan approval decisions,
utilizing a dataset containing various applicant attributes.

The initial phase involved comprehensive data preprocessing: handling
missing values, encoding categorical variables, and normalizing
numerical features. Exploratory data analysis was conducted to discern
patterns and relationships within the data, including the visualization.

Subsequently, dimensionality reduction techniques, such as Principal
Component Analysis (PCA), were applied to distill the most informative
features, thereby enhancing model performance and interpretability. The
modeling phase encompassed both supervised and unsupervised learning
approaches. Supervised methods included logistic regression, decision
trees, and k-nearest neighbors (KNN), while unsupervised techniques
involved clustering algorithms like k-prototypes and hierarchical
clustering based on Gower distance.

## Import Data

```{r echo=FALSE}
#load dplyr package for future use
library(dplyr)

# clear environment
rm(list = ls())

# set work directory
setwd("~/STAT 4601/Final Project")

# read dataset file
loandata <- read.csv("LoanApprovalPrediction.csv")

# examine data
str(loandata)
head(loandata)
```

The dataset consists of 598 cases and 13 attributes, which provide
information about applicants and their loan applications. The attributes
are as follows:

-   Loan_ID: A unique application ID assigned to each applicant.

-   Gender: The applicant's gender (Male or Female).

-   Married: The applicant's marital status (Married or Not Married).

-   Dependents: The number of people financially dependent on the
    applicant.

-   Education: The applicant’s education level (Graduate or Not
    Graduate).

-   Self_Employment: Indicates whether the applicant is self-employed.

-   ApplicantIncome: The total income of the applicant.

-   CoapplicantIncome: The total income of the co-applicant (if any).

-   LoanAmount: The loan amount requested by the applicant.

-   Loan_Amount_Term: The duration of the loan in months.

-   Credit_History: Indicates whether the applicant has a credit
    history.

-   Property_Area: The type of area where the applicant resides (Urban,
    Semi-Urban, or Rural).

-   Loan_Status: The outcome of the loan application (Approved or Not
    Approved).

In this report, we will apply machine learning techniques to predict
loan approvals based on applicant information. The target variable for
our analysis is "Loan_Status".

# Preprocessing and Data Analysis

## Missing Values and Duplicates

The dataset contains a total of 93 rows with a missing entry in at least
one of the columns. Through statistics, we can intuitively get the total
number of rows that have missing values and the number of missing values
in each variable.

```{r echo=FALSE}
# check if there are missing data entries
sum(rowSums(is.na(loandata)) > 0)

#check where the missing data are
data.frame(total_missing = colSums(is.na(loandata)), rate = paste0(round(colSums(is.na(loandata))*100/nrow(loandata), 2), "%"))
```

The total number of rows with missing values is 93. The columns which
contain the missing values are Dependents, LoanAmount, Loan_Amount_Term,
and Credit_History. To address missing values in dataset, employing mean
imputation for numeric variables and mode imputation for categorical
variables is a straightforward and effective approach. But before
applying those process, we need to check for Missing Completely At
Random (MCAR).

```{r echo=FALSE}
# Install and load the BaylorEdPsych package
library(misty)
library(mvnmle)
# Perform Little's MCAR test on your dataset
na.test(loandata)
```

Since for p-value = 0.133, this value is greater than the conventional
threshold of 0.05. We can conclued that fail to reject the null
hypothesis that the data are Missing Completely At Random (MCAR). This
suggests there's no strong evidence against the assumption that the
missingness in data is entirely random.

Thus for each numeric column with missing values, replace the NA entries
with the mean of the non-missing values in that column. And for
categorical variables, replace missing values with the mode (the most
frequent category).

```{r echo=FALSE}
# replace numeric missing values
for (col in c("Dependents", "LoanAmount", "Loan_Amount_Term")) {
  loandata[[col]][is.na(loandata[[col]])] <- round(mean(loandata[[col]], na.rm = TRUE))
}

# replace categoric missing values
loandata$Credit_History[is.na(loandata$Credit_History)] <- which.max(loandata$Credit_History)

# check if there are missing data entries
sum(rowSums(is.na(loandata)) > 0)
```

We've also assigned the proper datatype to each column and checked for
any duplicates but found none. Since there are no duplicate
applications, we can remove the ID variable as it will not be needed for
future analysis. We now have a total of 598 rows and 12 variables in our
dataset.

```{r echo=FALSE}
# check if there are any duplicates by substracting number of unique rows from number of total rows
nrow(loandata) - n_distinct(loandata)

# assign appropriate data type to each variable
loandata <- transform(loandata,
                    Gender = as.factor(Gender),
                    Married = as.factor(Married),
                    Education = as.factor(Education),
                    Self_Employed = as.factor(Self_Employed),
                    ApplicantIncome = as.numeric(ApplicantIncome),
                    LoanAmount = as.numeric(LoanAmount),
                    Loan_Amount_Term = as.numeric(Loan_Amount_Term),
                    Credit_History = as.factor(Credit_History),
                    Property_Area = as.factor(Property_Area),
                    Loan_Status = as.factor(Loan_Status))

# remove ID
loandata <- loandata[,-1]

# view data summary
str(loandata)
```

## Visualization

To begin our analysis, we examine the distribution of approved and
rejected loans in the dataset.

```{r echo=FALSE}
library(ggplot2)

# bar plot of frequency of loan satuses
ggplot(loandata, aes(x = Loan_Status, fill = Loan_Status)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +  # add count above each bar
  scale_fill_manual(values = c("Y" = "darkgreen", "N" = "darkred")) +
  labs(title = "Frequency of Loan Statuses", 
       x = "Loan Status", 
       y = "Frequency") +
  theme_minimal() +
  coord_cartesian(clip = 'off')

```

There is a clear imbalance in the number of approved versus denied loan
applications. Out of the total 598 cases, 411 applicants were granted a
loan, while 187 were denied.

We can also examine the distribution of all other attributes in our
dataset.

```{r echo=FALSE}
# bar graphs of categorical and discrete variables
for (i in colnames(loandata[,-c(6:9,12)])){
  
  plot.cat.attr <- ggplot(loandata, aes(x = !!sym(i))) +
    geom_bar(stat = "count", position = "dodge", fill = "#575366") +
    theme_minimal() +
    labs(title = paste("Distribution of", gsub("_", " ", i)),
         x = gsub("_", " ", i),
         y = "Frequency")
  print(plot.cat.attr)
  
}

# histograms of continuous variables
for (i in colnames(loandata[,c(6:9)])){
  
  plot.cts.attr <- ggplot(loandata, aes(x = !!sym(i))) +
    geom_histogram(fill = "#575366", color = "black", bins = 30) +
    labs(title = paste("Distribution of", gsub("_", " ", i)), 
         x = gsub("_", " ", i),
         y = "Frequency") +
    theme_minimal()
  
  print(plot.cts.attr)
  
}
```

The categorical variables with two levels exhibit significant imbalances
in their frequency distributions. Specifically, the dataset contains
more male applicants, married individuals, graduate students,
non-self-employed applicants, and individuals with a credit history.
This suggests that the majority of loan applicants share certain
demographic characteristics.

For the Property Area variable, which has three levels, the distribution
shows a similar number of rural and urban residents, while a larger
number of applicants live in semi-urban areas.

Regarding the Dependents variable, most applicants have no dependents,
while the number of applicants with one to three dependents is
significantly lower, indicating a right-skewed distribution.

Among the continuous variables, Applicant Income, Co-applicant Income,
and Loan Amount are heavily right-skewed, whereas Loan Amount Term
exhibits strong left skewness.

Given the imbalance within the categorical variables and the discrete
variable, rather than focusing solely on raw counts, we will analyze the
proportions of loan statuses within each of those variables to better
understand their relationship with loan outcomes.

```{r echo=FALSE}
#for loop to create bar graph for every attribute to see influence on loan status
for (i in setdiff(colnames(loandata[,-c(6:9)]), c("Loan_Status"))){
  
  #group by the factor and status of loan, count rows, add percentage and count
  attr_count <- loandata %>%
    group_by(!!sym(i), Loan_Status) %>%
    tally() %>%
    mutate(percentage = n / sum(n) * 100, TotalCount = sum(n))

  #bar graph with y=percentage and the count on top of each bar
  plotattr <- ggplot(attr_count, aes(x =!!sym(i) , y = percentage, 
                                     fill = Loan_Status)) +
    geom_bar(stat = "identity", col = "black") +
    geom_text(aes(label = round(percentage)),
              position = position_stack(vjust = 0.5), size = 2.5, 
              colour = "white") + 
    geom_text(aes(label = TotalCount), y=102.5, size = 2.5) +
    scale_fill_manual(values = c("Y" = "darkgreen", "N" = "darkred")) +
    theme_minimal() +
    labs(title = paste("Loan Status by", gsub("_", " ", i)),
         x = gsub("_", " ", i),
         y = "Percentage",
         fill = "Loan Status") 


  print(plotattr)
  
}
```

Among the examined features, credit history stands out as the strongest
predictor of loan approval. The "Credit History by Loan Status" plot
shows that applicants with a credit history are far more likely to be
approved, while those without one face a much higher likelihood of
rejection. This suggests that having a credit history is a key
requirement for banks to grant a loan.

Other attributes show weaker but still noticeable relationships with
loan status:

-   Marital Status: Married applicants had a slightly higher approval
    rate than unmarried ones.

-   Dependents: Applicants with two dependents saw slightly more
    approvals followed by those with no dependents, while those with one
    or three dependents had similarly lower approval rates.

-   Education: Graduates were approved at a slightly higher rate than
    non-graduates.

-   Property Area: Loan approvals were more common among applicants from
    semi-urban areas, while those from urban and rural areas experienced
    higher rejection rates in comparison.

Finally, self-employment status and gender have no apparent impact on
loan approval, suggesting that banks do not weigh these factors
significantly when making lending decisions.

For continuous variables, density plots were generated to compare the
distributions of approved and rejected applicants.

Before visualizing, zero values were excluded from co-applicant income,
as co-applicants are typically intended to provide financial support
when the primary applicant alone may not qualify for the loan. A value
of zero likely indicates that no co-applicant was involved, rather than
an actual absence of income. Since co-applicants are generally required
for low-income applicants or those with insufficient credit history,
removing these zeros ensures a more accurate representation in the
density plots.

```{r echo=FALSE}
#plot continuous variables for each loan ststua
for (i in setdiff(colnames(loandata[,c(6:9,12)]), c("Loan_Status"))) {
  
  #remove rows with coapplicant income of 0 for the coapplicant income plot
  if (i == "CoapplicantIncome"){
    ctsdata <- loandata %>%
      filter(CoapplicantIncome != 0)
  } else {
    ctsdata <- loandata
  }
  
  #create density plot with one function for each loan status
  plotdensity <- ggplot(ctsdata, aes(x = !!sym(i), fill = Loan_Status)) +
    geom_density(alpha = 0.6) +
    scale_fill_manual(values = c("Y" = "lightgreen", "N" = "darkred")) +
    labs(title = paste("Density Plot of", gsub("_", " ", i), "by Loan Status"),
         x = gsub("_", " ", i), 
         y = "Density", 
         fill = "Loan Status") +
    theme_minimal() +
    theme(legend.position = "top") +
    facet_wrap(~Loan_Status, scales = "free_y")
  
  print(plotdensity)
}

```

The density plots show that the distributions for approved and rejected
applicants are largely similar, with both peaking around the same
values. Although the curves are similar for both loan statuses, there is
a subtle but meaningful distinction. The density curve for approved
applicants appears slimmer, indicating a higher concentration around the
mode, whereas the rejected applicants exhibit greater variability. This
may suggest that banks approve loan applications when these continuous
variables fall within a preferred range.

## Encoding Categorical Variables

The dataset contains several binary variables, each with two categories.
These were encoded using binary representation, where values were
transformed into 0s and 1s. For the variable Property Area, which
includes three categories, ordinal encoding was applied by assigning
numeric values from 0 to 2.

Encoding the categorical variables with two levels into binary form
allows for the application of logistic Principal Component Analysis
(logistic PCA) later in the analysis. Logistic PCA is a dimensionality
reduction technique similar to standard PCA but specifically designed
for binary data.

The variable encodings are as follows:

-   Gender: 1 = Male, 0 = Female

-   Married: 1 = Married, 0 = Not Married

-   Education: 1 = Graduate, 0 = Not Graduate

-   Self_Employed: 1 = Self-employed, 0 = Not Self-employed

-   Property_Area: 0 = Urban, 1 = Semi-urban, 2 = Rural

-   Credit_History: 1 = Has a Credit History, 0 = No Credit History

-   Loan_Status: 1 = Approved, 0 = Denied

```{r echo=FALSE}
numeric <- c("Dependents", "ApplicantIncome", "CoapplicantIncome", "LoanAmount", "Loan_Amount_Term")
categoric <- c("Gender", "Married", "Education", "Self_Employed", "Property_Area", "Credit_History", "Loan_Status")
```

```{r echo=FALSE}
# converting categoricals to numeric
loanbinary <- loandata
loanbinary$Gender <- ifelse(loanbinary$Gender == "Male", 1,0)
loanbinary$Married <- ifelse(loanbinary$Married == "Yes", 1,0)
loanbinary$Education <- ifelse(loanbinary$Education == "Graduate", 1,0)
loanbinary$Self_Employed <- ifelse(loanbinary$Self_Employed == "Yes", 1,0)
loanbinary$Property_Area <- ifelse(loanbinary$Property_Area == "Urban", 0, ifelse(loanbinary$Property_Area == "Semiurban", 1, 2))
loanbinary$Loan_Status <- ifelse(loanbinary$Loan_Status == "Y", 1,0)

# make the categorical variables as factor
loanbinary[, categoric] <- lapply(loanbinary[, categoric], as.factor)

str(loanbinary)
```

## Correlation

To assess multicollinearity and redundancy among numeric variables, we
computed pairwise Pearson correlations between numerical variables
(Dependents, ApplicantIncome, CoapplicantIncome, LoanAmount,
Loan_Amount_Term).

```{r echo=FALSE, include=FALSE}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor) {
   usr <- par("usr"); on.exit(par(usr))
   par(usr = c(0, 1, 0, 1))
   r <- abs(cor(x, y))
   txt <- format(c(r, 0.123456789), digits=digits)[1]
   txt <- paste(prefix, txt, sep="")
   if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)*r
   text(0.5, 0.5, txt, cex = cex.cor)
}

     ## put histograms on the diagonal
panel.hist <- function(x, ...) {
   usr <- par("usr"); on.exit(par(usr))
   par(usr = c(usr[1:2], 0, 1.5) )
   h <- hist(x, plot = FALSE)
   breaks <- h$breaks; nB <- length(breaks)
   y <- h$counts; y <- y/max(y)
   rect(breaks[-nB], 0, breaks[-1], y, col="cyan", ...)
}
```

```{r echo=FALSE, warning=FALSE}
# Check coorelation
cormatrix <- cor(loanbinary[, numeric])
cormatrix
```

The strongest correlation is observed between ApplicantIncome and
LoanAmount (r ≈ 0.52), which is moderate and expected, as higher-income
applicants tend to request larger loans.\
All other variable pairs exhibit weak correlations (\|r\| \< 0.2),
suggesting that they provide largely independent information.

```{r echo=FALSE, warning=FALSE}
# print graph
pairs(loanbinary[, numeric], upper.panel = panel.cor, diag.panel = panel.hist)
```

Notably, CoapplicantIncome shows little to no correlation with the other
variables, including LoanAmount, indicating it may carry distinct
predictive value.

Overall, the low correlation among most numeric variables supports their
inclusion in modeling and dimensionality reduction without significant
concern for multicollinearity.

## Split data

When preparing data for building a classification, it's essential to
split the dataset into training and testing subsets to evaluate the
model's performance accurately.

The total amount of processed data is 598 obs., we divide it into
training data and test data in a ratio of 80% to 20% (commonly used).

```{r echo=FALSE}
# Split train data and test data
set.seed(123)

#load library to split test and train sets
library(caTools)

#split the data ensuring that the proportion of Loan_Status are preserved
#split the data so that around 80% of the data is in the training set and 20% is in the test set
split <- sample.split(loanbinary$Loan_Status, SplitRatio = 0.80)

#create train set
train.data <- subset(loanbinary, split == TRUE)

#create test set
test.data <- subset(loanbinary, split == FALSE)

table(train.data$Loan_Status)
table(test.data$Loan_Status)
```

The training set contains 479 observation in total with 329 approved
cases and 150 denied cases. The test set contains 119 total observations
in total with 82 approved cases and 37 denied cases.

## Undersampling

The original dataset exhibited a noticeable imbalance between approved
(Loan_Status = 1) and denied (Loan_Status = 0) loan applications, with
approvals comprising a significantly larger proportion. This imbalance
can bias supervised learning models toward the majority class, reducing
their ability to correctly identify minority-class instances.

To address this, we applied random undersampling to the majority class
in the training set, resulting in a balanced dataset with an equal
number of approved and denied cases.

Also, undersampling was not applied to the test data in order to
preserve its original class distribution, ensuring that model evaluation
reflects real-world performance on naturally imbalanced data.

```{r echo=FALSE}
#set seed for reproducibility
set.seed(123)

#store number of "No" loan state observations
num.mal <- nrow(train.data[train.data$Loan_Status == 0,])

# apply undersampling
balanced.data <- train.data %>%
  group_by(Loan_Status) %>%
  sample_n(size = num.mal) %>%
  ungroup()

# check again
table(balanced.data$Loan_Status)
```

## Dimension Reduction

To reduce the dimensionality of the dataset while preserving most of the
variance in the numeric features, we applied Principal Component
Analysis (PCA) exclusively to the numeric variables (Dependents,
ApplicantIncome, CoapplicantIncome, LoanAmount, and Loan_Amount_Term).

Standard PCA requires continuous, numerical input and does not offer
meaningfully interpretations for categorical values, even if encoded as
numbers, since the ordinal structure of such data may be arbitrary or
misleading.

```{r echo=FALSE}
# create subset of the data with only numeric variables
pca.numeric <- as.data.frame(balanced.data[, numeric])

# compute the principal components
pcamodel <- prcomp(pca.numeric, center = TRUE, scale = TRUE)

# summary of PCA
summary(pcamodel)
```

The resulting components show that the first principal component (PC1)
accounts for 29.9% of the total variance, and the first four components
together capture over 90.6% of the total variance. This justifies the
use of a reduced set of principal components in downstream analysis, as
they retain most of the informative variability of the numeric data
while minimizing redundancy.

```{r echo=FALSE}
# pca plot
plot(pcamodel, col = heat.colors(11))
```

We retained the first four principal components, which together capture
approximately 90.6% of the total variance in the numeric variables. This
strikes a balance between dimensionality reduction and information
retention, while excluding the fifth component, which contributes
minimal additional variance.

```{r echo=FALSE}
pcamodel$rotation[,1:4]
```

**PC1** = 0.71 x LoanAmount + 0.62 x ApplicantIncome + 0.24 x
(Dependents + CoapplicantIncome) + 0.09 x Loan_Amount_Term

-   PC1 captures the combined effect of the loan amount and the
    applicant's income.

**PC2** = - 0.87 x CoapplicantIncome + 0.41 x ApplicantIncome + 0.26 x
Loan_Amount_Term - 0.11 x LoanAmount

-   PC2 captures the effect of the co-applicant income with some
    contrast from the applicant income.

-   Hence, it it comparing how much the coapplicant earns against the
    salary of the applicant.

PC3 = -0.75 x Loan_Amount_Term + 0.63 x Dependents - 0.18 x
CoapplicantIncome - 0.11 x LoanAmount + 0.06 x ApplicantIncome

-   PC3 captures the effect of the loan amount term versus the number of
    dependents.

-   Hence, it is comparing how long the applicant will be granted the
    loan against how many people depend on them financially

PC4 = 0.74 x Dependents + 0.60 x Loan_Amount_Term - 0.29 x
ApplicantIncome -0.10 x LoanAmount + 0.06 x CoapplicantIncome

-   PC4 captures the combined effect of the number of dependents and the
    length of the loan.

-   PC3 is the contrast between the length of the loan and the number of
    dependents meanwhile PC4 is their merged effect.

For the binary variables, we applied Logistic Principal Component
Analysis (Logistic PCA) to perform dimensionality reduction on it.
Standard PCA is not suitable for binary data due to its assumption of
continuous variables. The principal components for the binary variables
will be referred to as PCB and the principal components for the
numerical variables will be referred to as PC.

Since Logistic PCA is only applicable to binary variables, we will
remove Property_Area because it is ternary. The target variable is also
not included.

```{r echo=FALSE}
library(logisticPCA)

# Categorical variables
cate <- categoric[!categoric %in% c("Property_Area", "Loan_Status")]

# get data into 1/0 binary numerical values
pca.categori <- as.data.frame(lapply(balanced.data[cate], function(x) as.numeric(as.character(x))))
```

```{r echo=FALSE, warning=FALSE}
# check for k and m
set.seed(123)
negloglikes <- cv.lpca((pca.categori), ks = 1:5, ms = 1:10)
plot(negloglikes)
```

We determined the optimal number of principal components is k=4 and the
saturation parameter is m=5 through cross-validation, selecting the
combination that minimized the negative log-likelihood.

```{r echo=FALSE}
(pcanummodel <- logisticPCA((pca.categori), k=4, m=5))

loadings <- pcanummodel$U
rownames(loadings) <- cate
```

A logistic PCA was performed on the binary categorical variables,
selecting k = 4 components. This configuration explained 94% of the
deviance, and shows a large amount of data variability is captured while
keeping the model simple. The choice balances the trade-off between
model complexity and explanation.

PCB1 = - 0.54 x Gender + 0.45 x (Self_Employed - [Education +
Credit_History]) - 0.32 x Married

-   All the variables have a high contribution

-   The highest contributor is the gender. Hence, PCB1 can be consider
    to mainly represent gender.

-   There is also significant contribution from self-employment status,
    educational status and presence of credit history, where
    self-employment status is contrasted with the other two.

PCB2 = - 0.73 x Married + 0.41 x Credit_History - 0.38 x Gender + 0.33 x
Education - 0.23 x Self_Employed

-   The main contributor is marital status with some comparison to the
    presence of a credit history.

-   Hence, PCB2 mainly represents the difference between the applicant's
    gender and their credit history.

PCB3 = - 0.67 x Education + 0.53 x Gender - 0.44 x Married - 0.23 x
Self_Employed + 0.13 x Credit_History

-   The two largest negative contributors are the education status and
    marital status.

-   The largest positive contributor is gender.

-   Thus, PCB3 is represented by the compared effect of educational and
    marital status against gender.

PCB4 = 0.71 x Self_Employed + 0.69 x Credit_History + 0.09 x (Gender -
Education) + 0.07 x Married

-   The dominating variables are the self-employment status and the
    presence of a credit history.

-   Therefore, PCB4 represents largely the combined effect of an
    applicant's self-employment status and their credit history.

```{r echo=FALSE}
# Combine numerical, categorical variables and Property_Area and target variable
train.pcadata <- cbind(pcamodel$x[, 1:4], pcanummodel$PCs, balanced.data[,c("Property_Area", "Loan_Status")])

# rename components from PCA for binary data
train.pcadata <- train.pcadata %>%
  rename(PCB1 = `1`,
         PCB2 = `2`,
         PCB3 = `3`,
         PCB4 = `4`)
```

After performing PCA, the numerical and categorical results were
combined into a single dataset along with the target variable and the
variable Property Area. This new dataset is the training set that will
be used for later analysis and modeling.

```{r echo=FALSE}
# convert the binary variables from the testing set to type numeric
test.data[cate] <- lapply(test.data[cate], function(x) as.numeric(as.character(x)))

# predict test data
test.pca1 <- predict(pcamodel, newdata = test.data[, numeric])
test.pca2 <- predict(pcanummodel, newdata = test.data[, cate])
test.pcadata <- cbind(test.pca1[, 1:4], test.pca2, test.data[, c("Property_Area", "Loan_Status")])

# rename components from PCA for binary data
test.pcadata <- test.pcadata %>%
  rename(PCB1 = `1`,
         PCB2 = `2`,
         PCB3 = `3`,
         PCB4 = `4`)
```

The same measures are applied to the test data sets to ensure
consistency. That is, the logistic and standard PCAs obtained from the
training set were applied to the test set.

# Machine Learning

## Unsupervised Learning

For this section, we will perform clustering analysis which is an
unsupervised learning method in order to group individuals with similar
loan applications into clusters. Two clustering methods will be used:
k-Prototypes and Gower Distance. Unsupervised learning differs from
supervised learning since it attempts to discover patterns in the data
without knowing the outcome. For this reason we will work with the raw
data as it is not required to use a training and tests set.

### k-Prototypes

The first clustering method we will use is k-Prototypes clustering which
can be applied to mixed data. K-prototypes is heavily influenced by the
majority classes since centroids are biased towards majority values.
Consequently, we've undersampled the loan dataset to balance the number
of approved and not approved loan applications in order to use the
k-prototype clustering method.

```{r echo=FALSE}
#set seed for reproducibility
set.seed(123)

#store number of "No" loan state observations
num.mal <- nrow(loanbinary[loanbinary$Loan_Status == 0,])

# apply undersampling
balanced.all.data <- loanbinary %>%
  group_by(Loan_Status) %>%
  sample_n(size = num.mal) %>%
  ungroup()

balanced.all.data <- as.data.frame(balanced.all.data)

# check again
table(balanced.all.data$Loan_Status)

# remove target variable
loan.kproto <- balanced.all.data %>%
  dplyr::select(!Loan_Status)
```

Once balanced, the data contained 187 observations for each the approved
loans and the denied loans. The target variable was then removed from
the dataset. To start, we first we need to determine the optimal number
of clusters. We can do so by using the standard scree test.

```{r echo=FALSE, include=FALSE}
#load required packages
library(clustMixType)
library(dplyr)
library(stats)

#set seed for reproducibility
set.seed(42)

#find the appropriate number of clusters using the standard scree test
Es <- numeric(10)
for(i in 1:10){
  loan.clust.kproto <- kproto(loan.kproto, k = i, nstart = 20)
  Es[i] <- loan.clust.kproto$tot.withinss
}
```

```{r, echo=FALSE}
#plot a scree plot of the Objective function
plot(1:10, Es, type = "b", 
     ylab = "Objective Function", 
     xlab = "# Clusters",
     main = "Scree Plot")
```

The scree plot of the objective function E reveals a subtansial drop
from k=1 to k=2 clusters. There is a slight decrease in the objective
function E from k=2 to k=5. Beyond k=5, the rate of decline becomes less
impactful, indicating minimized returns in model improvement. Using the
elbow method, the optimal number of clusters, where no stronger
improvements can be seen is k = 3.

Therefore, we select k=3 clusters for our analysis.

In applying the kproto function to our balanced data, we want to choose
nstart to be large enough to give us a better chance of finding global
optimum.

```{r echo=FALSE, include=FALSE}
#set seed for reproducibility
set.seed(42)

#k-prototype Clustering
loan.3proto <- kproto(loan.kproto, k = 3, nstart = 20)
```

```{r echo=FALSE}
# view clusters
loan.3proto
```

The results correctly indicate there are 5 numeric predictors and 6
categorical predictors. It also lists the sizes of each cluster which is
the number of points in a cluster and the within cluster error which is
the sum of squared distances between each point in a cluster and the
cluster’s centroid.

Cluster 1 has very few points and the smallest within cluster error.
Cluster 3 has a very large number of points and the largest within
cluster error.

```{r echo=FALSE}

#add cluster numbers into a dataset with loan status
loan.cluster <- balanced.all.data[,12, drop = FALSE]
loan.cluster$Cluster <- loan.3proto$cluster

#view distribution of loan status in clusters
table(loan.cluster$Cluster, loan.cluster$Loan_Status)
prop.table(table(loan.cluster$Cluster, loan.cluster$Loan_Status), margin = 1)

```

The distribution of loan status within each cluster is alarming.
Although there is a dominating loan status within each cluster, the
split is almost 50-50.

If we consider the majority class within each cluster, we can assign
cluster 1 and 2 to denied loan applications and cluster 3 to approved
loan applications.

Let’s view the summary and the visualization for more details on how
much proportion of each feature the clusters represents:

```{r echo=FALSE}
#view summary
summary(loan.3proto)
```

```{r echo=FALSE}
#2 by 2 grid for plots
par(mfrow=c(2,2))

#create plots to view the distributions
clustMixType::clprofiles(object = loan.3proto, x = balanced.all.data)
```

In the plots, cluster 1 is represented in green, cluster 2 in yellow and
cluster 3 in purple.

**Cluster 1 (1 approved loan, 2 denied loan):**

This cluster has a size of three points. It consists entirely of
applicants that are male, married, graduates, and have no co-applicant
income. The number of dependents vary and the property area in which
they live is evenly distributed. Comapred to the other clusters, they
have the highest requested loan amount and the lowest loan amount term
with a narrower range. In comparing this cluster's proportional
distribution against the proportional distribution of the other two
clusters, it has the most self-employed applicants and the most
applicants with no credit history.

Many of the attributes are skewed due to this cluster containing a very
small number of points. As a result, this cluster dominates many of the
attributes. Considering this, for the next two clusters, we will compare
them with each other and exclude cluster 1.

**Cluster 2 (60 approved loan, 75 denied loan) and Cluster 3 (126
approved loan, 110 denied loan):**

Cluster has a size of 135 points and Cluster 3 has a size of 236 points.
Proportionally, Cluster 2 contains more individuals that are female,
unmarried, self-employed, and slightly more those with a credit history
and that are graduates than Cluster 3. On average, applicants in Cluster
2 have around 0 dependents and those in Cluster 3 have around 1.
Compared to cluster 3, Cluster 2 has on average applicants with a higher
income, lower co-applicant income, requested higher loan amount and
within a wider range, and a slightly higher loan term with a narrower
range. Cluster 2 had applicants that mostly live in urban areas and
Cluster 3 has applicants that mostly live in semi-urban areas.

### Gower Distance

The next method we will use is Gower distance. It can be used with class
imbalance datasets and so we will use the processed loan data.

```{r echo=FALSE}
#load required packages
library(cluster)
library(Rtsne)
library(ggplot2)
library(prcr)

#calculate gower distance
#remove target before and specify the types
gower_dist <- daisy(loanbinary[, -12],
                    metric = "gower",
                    type = list(asymm = c("Gender", "Married", "Education",
                                          "Self_Employed", "Credit_History"),
                                factor = c("Property_Area"),
                                logratio = c("ApplicantIncome", "Dependents",
                                             "CoapplicantIncome", "LoanAmount"),
                                integer = c("Loan_Amount_Term")))

summary(gower_dist)
```

The plots in the visualization section show that the numeric variables
Applicant Income, Dependents, coapplicant Income and Loan Amount are
right skewed, so a log transformation will be applied to these variables
when calculating the Gower distance. Additionally, we need to specify
the variable types for the categorical variables. The binary variables
are all asymmetric as was seen in the visualization section of this
report, and so will be entered as asymmetric binary variables. As for
property area, it will be considered as a factor. Since we want to
exclude the target variable, Loan_Status, from the Gower distance
calculation, we have removed it from the analysis.

We can display the most similar and dissimilar pairs of points to verify
for any odd results.

```{r echo=FALSE}
#format as matrix
gower_mat <- as.matrix(gower_dist)

#output most similar pair
loanbinary[which(gower_mat == min(gower_mat[gower_mat != min(gower_mat)]),
        arr.ind = TRUE)[1, ], ]
```

The most similar pair of points have many equal attributes or attributes
with values that are close in range. Their attributes are also those
that are most common in the data. Both are applicants with an approved
loan status.

```{r echo=FALSE}
#output most dissimilar pair
loanbinary[which(gower_mat == max(gower_mat[gower_mat != max(gower_mat)]),
        arr.ind = TRUE)[1, ], ]
```

As for the dissimilar pair all their attributes are opposing and have a
great difference between their values. However, both are applicants with
a denied loan status.

```{r echo=FALSE}
#calculate silhouette width for many k using PAM
sil_width <- c(NA)

for(i in 2:20){
  pam_fit <- pam(gower_dist,
                 diss = TRUE,
                 k = i)
  
  sil_width[i] <- pam_fit$silinfo$avg.width
}

#plot sihouette width (higher is better)
plot(1:20, sil_width,
     xlab = "Number of clusters",
     ylab = "Silhouette Width")
lines(1:20, sil_width)
```

We want to choose the number of cluster with the highest silouette
width. Using the elbow method, we will select k = 3.

Let's view the summary and the medoids:

```{r}
pam_fit <- pam(gower_dist, diss = TRUE, k = 3)

pam_results <- loanbinary %>%
  mutate(cluster = pam_fit$clustering) %>%
  group_by(cluster) %>%
  do(the_summary = summary(.))

pam_results$the_summary
```

Based on the cluster summaries, the model did not produce strongly
distinct groupings with respect to loan status. They have instead
perfectly clustered by property area.

Examining the medoids, which are the most representative data points
within each cluster, provides some insight. The medoids are as follows:

```{r}
#view medoids
loanbinary[pam_fit$medoids, ]
```

Although many binary variables (Gender, Married, Education,
Self-Employed, and Credit History) are asymmetric and do not vary across
medoids, some distinctions emerge in the other features.

The most notable differences between the medoids lie in the number of
dependents, the range of applicant and co-applicant incomes, and the
property area. The medoid for Cluster 1 has 2 dependents, the lowest
applicant and co-applicant incomes among the three, and an urban
property area. The medoid for Cluster 3 has 0 dependents, the highest
applicant and co-applicant incomes, and a semi-urban property area.
Meanwhile, the medoid for Cluster 2 falls between the other two in terms
of income, has 1 dependent, and is associated with a rural property
area.

```{r}
tsne_obj <- Rtsne(gower_dist, is_distance = TRUE)

tsne_data <- tsne_obj$Y %>%
  data.frame() %>%
  setNames(c("X", "Y")) %>%
  mutate(cluster = factor(pam_fit$clustering),
         name = loanbinary$Loan_Status)

ggplot(aes(x = X, y = Y), data = tsne_data) +
  geom_point(aes(color = cluster))
```

The clusters are very well-separated, with no overlapping points.
However this clustering has reflected more upon the property area of the
applicants rather than their loan status. Cluster 1 and 3 which
represent the approved loans are further south on the graph than Cluster
2 which represents the denied loans.

Hence, the Gower distance clustering method did not effectively separate
the clusters based on loan status but instead their property area. The
medoids did not differ greatly but did identify two clusters that are
primarily associated with approved loans and one for not approved.

## Supervised Learning

The goal of supervised learning in this project is to develop predictive
models that classify loan applications as approved or denied based on
applicant attributes. We compare three algorithms, logistic regression,
decision tree, and random forest using metrics such as accuracy,
sensitivity, specificity, and area under the ROC curve. This comparison
highlights trade‑offs between predictive performance and
interpretability to inform model selection for real‑world deployment.

### Logistic

We begin with a logistic regression model to estimate the probability of
loan approval as a function of all predictor variables.

```{r}
# Fit the logistic regression model using all available predictors
logit_model <- glm(Loan_Status ~ ., data = train.pcadata, family = binomial)

# Display summary of the fitted model to inspect coefficients and significance
summary(logit_model)
```

The intercept is highly significant, equal to the sample mean for the
urban property area and the sample mean of the response when the PCs
(numeric variables) are equal to 0.

Out of all the predictors, we have 4 that are significant; PCB1, PCB2,
PCB4, and semi-urban property area. This model suggest that the PCs have
very little statistical significance in relation to the response. Hence,
PCA on numeric variables did not improve predictive ability.

PCB1: Dominated by Gender, Self_Employed, Education, and Credit_History
Estimate: -0.210327

Interpretation: A higher score on PCB1 (e.g., being male, self-employed,
less educated, or lacking credit history) is associated with a decrease
in the log-odds of loan approval.

PCB2: Primarily influenced by Marital Status and Credit_History
Estimate: 0.119837

Interpretation: A higher PCB2 score (e.g., being unmarried with a good
credit history) correlates with an increase in the log-odds of loan
approval.

PCB4: Driven by Self_Employed and Credit_History Estimate: 0.215232

Interpretation: Individuals who are self-employed and have a credit
history (higher PCB4 scores) are more likely to have their loans
approved.

All three principal‐component predictors are highly statistically
significant in the loan‐approval model at the α=0.001 level. This robust
evidence confirms that the underlying factors:gender, self‑employment,
education, marital status, and credit history which captured by these
components exert a strong and reliable influence on the log‑odds of loan
approval.

```{r}
# Predict probabilities and class labels on the test set
test_probs <- predict(logit_model, newdata = test.pcadata, type = "response")
test_pred  <- ifelse(test_probs > 0.5, 1, 0)
```

```{r}
# confusion matrix
library(caret)
conf_matrix <- confusionMatrix(as.factor(test_pred), as.factor(test.pcadata$Loan_Status))
print(conf_matrix)
```

True Positives (TP) = 24

True Negatives (TN) = 64

False Positives (FP) = 18

False Negatives (FN) = 13

```{r}
# calculate AUC result
library(pROC)
auc_value <- auc(test.pcadata$Loan_Status, test_probs)

# Create a data frame with the performance metrics
logperformance <- data.frame(
  Value = c(conf_matrix$overall['Accuracy'],
            conf_matrix$byClass['Recall'], 
            conf_matrix$byClass['Precision'], 
            conf_matrix$byClass['F1'], 
            auc_value)
)

rownames(logperformance) <- c("Accuracy", "Recall", "Precision", "F1", "AUC")

# Print the summary table
print(logperformance)
```

**Fair Recall (65%)**

We’re catching only about 65% of the actual approvals, whereas 35% of
valid “approve” cases are slipping through as false negatives.

**Less Than Fair Precision (57%)**

When the model says “approve,” it’s correct roughly 57% of the time,
meaning 43% of those flagged get rejected in reality. This result is not
ideal as it is just above random.

**Fair F1 (61%)**

Precision and recall are slightly distanced, yielding an F1 of 61%.
There is some bias toward one error type.

**Moderate Accuracy (74%)**

Good overall, but reflects the underlying class imbalance (around 74%
approvals).

**Good AUC (80%)**

The model has good discriminative power, but there’s room for
improvement. Ideally an AUC of above 0.80 has stronger separation.

### Tree Classification

Next, we train a decision tree classifier using the rpart package.

```{r echo=FALSE}
# apply classification tree rpart() method
library(rpart)

# Build the tree model using the 6 principal components to predict Class
treemodel <- rpart(Loan_Status ~ ., data = train.pcadata, method = "class", minbucket = 10)

x <- summary(treemodel)
```

The decision‑tree model distilled the 300 loan records into a
seven‑split structure that is both interpretable and reasonably
accurate:

Key drivers: Two transformed features—PCB4 and PCB2—account for almost
80% of the total split improvement. A very low PCB4 (\< –3.55) alone
flags a loan as almost certainly bad (node2, error ≈3%), while an
extremely low PCB2 (\< –5.13) performs the same gate‑keeping role for
the rest of the portfolio (node6).

Refinements: Applicants who clear those two hurdles are sorted by PC4,
Property_Area, and, to a lesser extent, PC3/PC1. These secondary splits
raise the posterior probability of a good loan (class1) from roughly 60%
at node3 to nearly 90% in the “sweet‑spot” leaf.

Model size vs reliability Cross‑validation points to the subtree with
CP=0.02 as the sweet spot: it trims three low‑value splits without
sacrificing predictive power, holding the 10‑fold mis‑classification
rate at ≈33%. Keeping additional splits (CP\<0.02) yields no measurable
gain.

Performance perspective: A 33% error rate is in line with the 25–40%
leaf‑level losses shown in the print‑out, suggesting the tree
generalises about as well as it fits. Compared with our earlier k‑NN and
logistic models, the tree trades a few points of accuracy for full
transparency in how each decision is reached.

```{r echo=FALSE}
plot(treemodel, main = "Classification Tree using rpart()", margin = 0.1)
text(treemodel, use.n = T, all = T, cex = 1)
```

The pruned classification tree shows that loans with very negative PCB4
values (\< –3.55), or if that hurdle is cleared—very negative PCB2
values (\< –5.13), are overwhelmingly likely to default, whereas the
remaining applicants are progressively sorted by property area and
mid‑range PC4/PC3 thresholds into branches whose default risk steadily
declines, ending in a semi‑urban subgroup with moderate component scores
where roughly 9 out of 10 loans succeed, thereby cleanly segmenting the
entire portfolio into high, medium, and low risk tiers with just seven
easy to explain rules.

```{r echo=FALSE}
# predict the test data
treepred <- predict(treemodel, newdata = test.pcadata, type = "class")
summary(treepred)

# compare the predict result with real data
table(as.integer(as.character(unname(treepred)))-as.integer(as.character(test.pcadata$Loan_Status)))
```

This output is showing, for each test case, the difference:
predicted_class − actual_class –1 (count=24) Predicted 0 (deny) but
actual1 (approve) =\> false negatives (we missed 24 real approvals).

0 (count=79) Predicted = actual =\> correct predictions (79 in total).

1 (count=16) Predicted 1 (approve) but actual 0 (deny) =\> false
positives (we over‑approved 16 denials).

```{r echo=FALSE}
# confusion matrix
library(caret)
confusion.matrix <- confusionMatrix(treepred, as.factor(test.pcadata$Loan_Status))
print(confusion.matrix)
```

Confusion Matrix

True negatives (correctly predicted approvals) = 58

True positives (correctly predicted denials) = 21

False positives (predicted deny but actually approve) = 24

False negatives (predicted approve but actually deny) = 16

With this test set the tree gets 66% overall accuracy, but because
always guessing the majority class would already give 69%, the gain is
statistically insignificant (p≈0.76) and agreement beyond chance remains
only fair (k≈0.26); it manages to catch 57% of the truly bad loans
(sensitivity) while correctly clearing 71% of the good ones
(specificity), yet its low precision of 47% means that more than half
the applicants it flags as “bad” are actually good, whereas its negative
predictive value of 78% shows that borrowers it labels “good” are indeed
safe nearly four times out of five, producing a balanced accuracy of
0.64 that signals moderate,but not compelling discriminatory power.

```{r echo=FALSE}
# calculate AUC result
library(pROC)
aucresult <- auc(roc(as.factor(test.pcadata$Loan_Status), as.integer(as.character(unname(treepred)))))


# Create a data frame with the performance metrics
treeperformance <- data.frame(
  Value = c(confusion.matrix$overall['Accuracy'],
            confusion.matrix$byClass['Recall'], 
            confusion.matrix$byClass['Precision'], 
            confusion.matrix$byClass['F1'], 
            aucresult)
)

rownames(treeperformance) <- c("Accuracy", "Recall", "Precision", "F1", "AUC")

# Print the summary table
print(treeperformance)
```

Not beating the naive baseline: Because only 31% of applicants are
actually bad, a trivial rule that rubber stamps everyone as good hits
69% accuracy,higher than the tree’s 66%. The McNemar and NIR p‑values
confirm the gap is not statistically significant.

Moderate recall, low precision: Recall 0.57 says the model intercepts
only a bit more than half the truly risky loans. Precision 0.47 says
that, of the applicants it does decline, more than half would have
repaid—a costly loss of business.

Skew toward false alarms: The model is more willing to shout “risk” than
to miss danger, but it still misses 16 bad loans and simultaneously
blocks 24 good ones in this 119 loan test sample. That is a 1:1.5
trade‑off between avoided defaults and lost revenue.

Discrimination is only “fair.” An AUC of 0.64 and balanced accuracy of
0.64 place the tree in the lower middle tier of credit risk models.

### KNN

To complement our tree‐based and regression approaches, we apply a
k‑nearest neighbors (kNN) classifier, which predicts loan approval by
finding the k most similar past applications in feature space.

```{r echo=FALSE}
library(class)

#training set without target variable
train <- train.pcadata[, 1:9]

#testing set without target variable
test <- test.pcadata[, 1:9]

#load library for knn
library(class)

#create vector that will store the accuracies
error.k <- c()

#calculate out of sample error the best k for odd k from 1 to 50
for (k in seq(1, 40, by = 2)) {
  class.knn <- knn(train = train, 
                        test = test, 
                        cl = train.pcadata$Loan_Status, 
                        k = k)
  
  Error <- mean(class.knn != test.pcadata$Loan_Status) 
  error.k <- c(error.k, 1 - Error) #store accuracy
  print(paste('k =', k, 'Accuracy =', 1 - Error)) 
}
```

At k=27, the k‑NN classifier reaches its peak accuracy of ≈84%, the
highest of all tested values, placing it firmly in the optimal
bias–variance regime.

```{r echo=FALSE}
#plot accuracies for different values of k
plot(seq(1, 40, by = 2), error.k, type = "b", xlab = "k", ylab = "Accuracy",
     main = "k-NN Accuracy for different k")
```

k≤7(high variance) – Accuracy hovers around 0.69‑0.72 and even dips at
k=5, showing that a very small neighborhood overfitts to local noise.

k=9‑17(stabilizing) – As it let each query look at more neighbors,
accuracy climbs sharply into the 0.78‑0.82 band; the model is now
smoothing out idiosyncratic points without yet blurring genuine class
boundaries.

k=19‑31(plateau) – The curve flattens at its high point: 0.82‑0.84,
peaking at k=27. Any value in this strip delivers virtually the same
performance, so pick the lower end k=19 if we need faster predictions
and the upper end k=27 if we need the single best score.

k≥33(rising bias) – Accuracy tilts down slightly toward 0.81 as the
neighborhood grows too large and the classifier starts averaging across
heterogeneous regions.

```{r echo=FALSE}
#preform knn for k=28
(knnpred <- knn(train = train, test = test, cl = train.pcadata$Loan_Status,  k=21))
```

```{r}
#load library for confusion matrix
library(caret)

#create confusion matrix
(knn.conf.mat <- confusionMatrix(knnpred, test.pcadata$Loan_Status))
```

True Positives (TP=18) – Loans the model flags as bad and that are
indeed bad.

False Positives (FP=4) – Loans the model flags as bad but are actually
good (costly lost business).

False Negatives (FN=19) – Bad loans the model mistakenly approves
(credit‑loss risk).

True Negatives (TN=78) – Good loans correctly approved.

This k‑NN setup is exceptional at sparing good customers, only 4 false
rejections—and reliable when it does decline someone. Its one
shortcoming is that it still misses roughly half of the truly bad loans,
so if default avoidance is paramount we can layer in a second check or
adjust the decision threshold to boost recall. Overall, k‑NN offers a
strong trade off for customer‑friendly credit decisions, with
statistically significant lift over a naïve baseline.

```{r echo=FALSE}
library(pROC)
auc_result <- auc(roc(as.factor(test.pcadata$Loan_Status), 
                      as.integer(as.character(unname(knnpred)))))


# Create a data frame with the performance metrics
knnperformance <- data.frame(
  Value = c(knn.conf.mat$overall['Accuracy'],
            knn.conf.mat$byClass['Recall'], 
            knn.conf.mat$byClass['Precision'], 
            knn.conf.mat$byClass['F1'], 
            auc_result)
)

rownames(knnperformance) <- c("Accuracy", "Recall", "Precision", "F1", "AUC")
knnperformance
```

The k‑NN model delivers a 0.81 overall accuracy and an AUC of 0.72,
meaning it separates good versus bad loans substantially better than
chance. Its precision of 0.82 tells us that when the model flags someone
as “bad,” it is correct more than four‑fifths of the time, so we avoid
needless rejections. At the same time, its recall of 0.49 means it
catches roughly half of all actual defaults, and the resulting F1 score
of 0.61 reflects a solid but improvable balance between capturing risky
loans and sparing good ones.

```{r echo=FALSE}
comparetable <- data.frame(
  knn = c(knnperformance[,1]),
  logistic = c(logperformance[,1]), 
  tree = c(treeperformance[,1])
  )

rownames(comparetable) <- c("Accuracy", "Recall", "Precision", "F1", "AUC")

print(comparetable)
```

Across all five metrics the rpart tree clearly under‑performs, so we can
set it aside. The real contest is between logistic regression and k‑NN:

k‑NN

Accuracy=0.81 (best)

Precision=0.82 (best)

Recall=0.49 (lowest)

F1=0.61

AUC=0.72

k‑NN excels at sparing good customers when it says “decline” it’s right
82% of the time—and delivers the highest overall hit rate, but it only
catches half of the bad loans and its discrimination (AUC) is moderate.

Logistic regression

Accuracy=0.74

Precision=0.57

Recall=0.65 (best)

F1=0.61

AUC=0.80 (best)

Logistic regression offers the strongest ranking ability (AUC≈0.80) and
the highest recall, so it intercepts nearly two thirds of defaults, even
though it sacrifices some precision and overall accuracy.

Because both k‑NN and logistic achieve essentially the same F1 (around
0.61), the choice hinges on business priorities:

If avoiding false declines is paramount (retaining good customers), k‑NN
is our top pick.

If catching as many defaults as possible is critical which means
minimising loan losses, then logistic regression is preferable.

For a production system, we recommend logistic regression as the
default, its superior AUC and recall give us the best risk separation,
while tuning its probability threshold or combining it with a precision
which is focused second check to control how many good customers we turn
away.

## Conclusion

In this project, we developed two unsupervided learning models;
k-Prototypes and Gower Distance, and three supervised learning models;
logistic regression, decision tree, and k‑nearest neighbors (k‑NN).
These were done with the goal to predict loan approval.

First we preprocessed the data and applied two dimension reduction
methods. The first was the standard PCA for the continuous variable. We
selected the first 4 components that covered around 90% of the
variability. The second method was logistic PCA for binary variables. We
selected the first 4 components from this model that also covered around
90% of the variance in the data.

The unsupervised learning models were constructed using the processed
data for Gower distance and the balanced processed data for
k-Prototypes. The clusters created by both models did not succeed with
separating very well by loan status. The k-Prototypes model had a number
of both approved and rejected loan that were very close in range and
created a cluster of only 3 points. The Gower Distance also did not
succeed very well by clustering according to property area rather than
loan status.

The supervised models were fitted using the balanced training set and
the test set.

Data preprocessing & PCA

Raw borrower attributes: income, loan amount, term and dependents were
transformed via principal components (PC1–PC4) to capture the main
sources of variance.

Model training & evaluation

We fit each model on a 70/30 train/test split, measuring accuracy,
recall (sensitivity), precision, F₁, and AUC on the hold out set.

Model selection & business trade‑off

The decision tree is easiest to explain but underperforms materially.

k‑NN offers the highest accuracy and spares good customers which only
has 4 false alarms, making it ideal where minimizing customer friction
is critical.

Logistic regression strikes the best balance of recall and ranking
powerkey when avoiding loan default carries the greatest cost.

Overall, our analysis suggests that logistic regression provides the
best basis for producing loan approval scorecards, while k-NN can serve
as an accuracy focused enhancement model in multistage decision-making
processes.

data visualization：Selina Salameh

data reduction：Lening Gao

dimension reduction：Lening Gao (interpretions of components by both
Lening Gao and Selina Salameh)

unsupervised learning：Selina Salameh

supervised learning：Xiwei Zhang
